Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-09-30 manuscript/abstract.tex
Update key findings, fix figure placement, specify test set size

Key points:

\begin{itemize}
\item{Phenotypic measurements of antigenic advance outperform genetic estimates from epitope mutations in long-term forecasts}
\item{DMS measurements enable more precise identification of mutationally intolerant sites to replace previous sites based on all non-epitope regions}
\item{Phylogenetic fitness metrics outperform and appear to be mutually exclusive of mechanistic metrics}
\item{Distance-based model target performs as well or better than clade-based targets and enables future forecasting efforts where phylogenies are not feasible}
\item{We provide (the first?) open source, extensible forecasting framework for influenza}
\end{itemize}

>>>>

\section*{Key findings}

\begin{itemize}
\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
\item{Experimental measurements of antigenic drift based on HI assays are more robust and predictive of viral success than epitope mutations}
\item{The combination of LBI and non-epitope mutation metrics outperforms all experimentally-informed metrics}
\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
\end{itemize}

\section*{Open questions}

\begin{itemize}
\item{How should forecast uncertainty be calculated?}
\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance?}
\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?}
\item{How does our model performance compare with the original clade-based models?}
\item{Should we report both validation and test results? Should we report them together or separate?}
\item{Should we report difference from naive as a primary statistic? As naive - model or as model - naive?}
\end{itemize}

### 2019-09-30 manuscript/main.tex
Update key findings, fix figure placement, specify test set size

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
  \caption{
    Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.
    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) with the mean $\pm$ standard deviation in the top-left corner of each panel.
    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
    Models outperform the naive model when the model's distance to the future is less than the naive model.
    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
  \end{center}
\end{figure*}

@@

\begin{itemize}
\item{Our model accurate forecasts evolution of simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics}
\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
\item{Epitope mutations are an inadequate predictor of viral success compared to other antigenic or phylogenetic metrics}
\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
\end{itemize}

@@

For model training and validation, we selected 13,568 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1992 and October 1, 2015.
To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
We excluded all egg-passaged samples and all samples with ambiguous year, month, and day annotations.
We prioritized samples with more available HI titer measurements.
For model testing, we selected an additional XX HA sequences that were sampled between October 1, 2015 and April 1, 2019.
We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.

>>>>

\begin{figure*}[ht!]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
  \caption{
    Model a) coefficients and b) accuracy for simulated populations of A/H3N2-like viruses.
    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) with the mean $\pm$ standard deviation in the top-left corner of each panel.
    Distances to the future in amino acids are shown per validation timepoint for each individual model (black) in the context of the observed distance between timepoints from the naive model (grey).
    Models outperform the naive model when the model's distance to the future is less than the naive model.
    The mean $\pm$ standard deviation of amino acids per timepoint are shown in the top-left of each panel for both the model and the naive model.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
  \end{center}
\end{figure*}

@@

Key findings:

\begin{itemize}
\item{Our model accurately forecasts clade frequency trajectories and identifies optimal potential vaccine strains for both simulated and natural populations by estimating the sequence composition of future populations and without relying on clade-based model targets}
\item{Experimental measurements of antigenic drift based on HI assays are more predictive of viral success than epitope mutations}
\item{The combination of sequence-only metrics outperforms all experimentally-informed metrics}
\item{Deleterious mutations contribute more to seasonal influenza evolution than has been as widely appreciated}
\item{DMS measurements appear to be too background-specific to be predictive for global viral success}
\item{Further efforts to understand the declining efficacy of HI assays and to replace these with FRA or antigenic escape assays could improve models in the future}
\item{Our model live forecasts in nextstrain.org will aid in year-round surveillance of influenza evolutionary patterns and allow us to continuously evaluate model performance relative to recent observations}
\item{Our model is the first of its kind to be released as an open source framework that can be inspected and extended by others}
\item{Immediate next steps to improve influenza models under this framework include the integration of geographic information and antigenic escape assay data}
\item{Our distance-based model targets and easy definition of fitness metrics through tidy data frames paves the way for future forecasting efforts with pathogens that cannot be analyzed by standard phylogenetic methods (e.g., highly recombinant viruses and organisms with larger genomes like bacteria and fungi)}
\end{itemize}

@@

For model training and validation, we selected 13,568 HA sequences $\geq$900 nucleotides that were sampled between October 1, 1992 and October 1, 2015.
To account for known variation in sequence availability by region, we subsampled the selected sequences to a representative set of 90 viruses per month with even sampling across 10 global regions including Africa, Europe, North America, China, South Asia, Japan and Korea, Oceania, South America, Southeast Asia, and West Asia.
We excluded all egg-passaged samples and all samples with ambiguous year, month, and day annotations.
We prioritized samples with more available HI titer measurements.
For model testing, we selected an additional 20,609 HA sequences corresponding to 90 viruses per month sampled between October 1, 2015 and April 1, 2019.
This increased number of sequences corresponds with the trend of increased sample sequencing and deposition in the last decade.
We used these test sequences to evaluate the out-of-sample error of fixed model parameters learned during training and validation.
